---
title: "Forest Fires NE Portugal"
author: "Shaurya Srivastava, Ishas Kekre, Anjali Viramgama"
date: "5/18/2020"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'docs/index.html')) })
output:
  html_document:
    theme: united
    highlight: breezedark
    number_sections: true
---

CMSC320 Final Project

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(tidyverse)
library(broom)
library(gapminder)
library(stringr)
```
 
# Introduction

# Forest Fires Information

# Dataset Used

The dataset used is from http://archive.ics.uci.edu/ml/datasets/Forest+Fires. It contains data about forest fires that have occurred at Northeast Portugal.

# Understanding the Dataset

# Data Preparation

## Load the Dataset

```{r}
csv_file <- "forestfires.csv"
forest_fires <- read.csv(csv_file, stringsAsFactors = FALSE)
forest_fires %>% head()
```

## Modify Dataset as Needed 

Normally, this means dropping any columns that are unnecessary and adding columns that will be useful for further data analysis.

### Add a column with the log of the areas. Drop the day column. Convert the month abbreviations to numbers

```{r}
forest_fires_mod <- forest_fires %>% 
  mutate(log_area = log(forest_fires$area + 1)) %>%
  select(-day) 
forest_fires_mod$month <- sapply(forest_fires_mod$month,function(x) grep(paste("(?i)",x,sep=""),month.abb))
```

### Add a column for the season

The season column is based on the month column. 

```{r}
forest_fires_mod <- forest_fires_mod %>% 
  mutate(season="winter")
forest_fires_mod$season[forest_fires_mod$month>2 & forest_fires_mod$month<6] <- "spring"
forest_fires_mod$season[forest_fires_mod$month>5 & forest_fires_mod$month<9] <- "summer"
forest_fires_mod$season[forest_fires_mod$month>8 & forest_fires_mod$month<12] <- "fall"
forest_fires_mod$season <- as.factor(forest_fires_mod$season)
```

### Add a column for the burn area level

There are two burn area levels, small and large. Burn area has a small level when log(area) <= 2 and a large level when log(area) > 2.

```{r}
forest_fires_mod <- forest_fires_mod %>% 
  mutate(burn_area_level="small")
forest_fires_mod$burn_area_level[forest_fires_mod$log_area>2] <- "large"
forest_fires_mod$burn_area_level <- as.factor(forest_fires_mod$burn_area_level)
```

### Display the first few rows of the new modified dataframe.

```{r} 
forest_fires_mod %>% head()
```

# Exploratory Data Analysis

``` {r hist}
forest_fires_mod %>%
  ggplot(mapping=aes(x=month),  breaks=100) +
    ggtitle("                                                   Histogram of Months") +
    geom_histogram(aes(y=..density..),binwidth=1.5) +
    geom_density(alpha=.2, fill="#FF6666") +
    scale_x_continuous(name ="Months", 
                    breaks=seq(1,12,1))
```

``` {r pie}
t = nrow(forest_fires_mod)
s = nrow(forest_fires_mod %>% filter(season == 'summer'))
sp = nrow(forest_fires_mod %>% filter(season == 'spring'))
f = nrow(forest_fires_mod %>% filter(season == 'fall'))
w = nrow(forest_fires_mod %>% filter(season == 'winter'))
seasons <- data.frame(
  group = c("Summer", "Spring", "Fall", "Winter"),
  value = c(s/t, sp/t, f/t,w/t)
  )
seasons %>%
  ggplot(aes(x="", y=value, fill=group)) + geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) + geom_text(aes(label = paste0(round(value*100), "%")), position = position_stack(vjust = 0.5)) +
  labs(x = NULL, y = NULL, fill = NULL, title = "                             Seasonal Fires") +
  theme_classic() + theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())
```

``` {r map}

forest_fires_mod %>% 
  ggplot(aes(x=X, y=Y) ) +
  ggtitle("                                           Forest Fire Map") +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  scale_x_continuous(name ="X-coord", 
                    breaks=seq(1,9,1)) +
  scale_y_continuous(name ="Y-coord", 
                    breaks=seq(2,9,1))
```

# Hypothesis Testing

# H0: p = 0.25 Ha: p > 0.25 for fall
```{r fall}
fall_sample <- forest_fires_mod %>% filter(season == 'fall')
xbar = nrow(fall_sample)/nrow(forest_fires_mod)
u = 0.25
sigma = 0.25*(1-0.25)/nrow(forest_fires_mod)
z = (xbar-u)/(sqrt(sigma))
pval = pnorm(z, lower.tail=FALSE)
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", sqrt(sigma)), quote = FALSE)
print(paste0("z-score: ", z), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# H0: u = 20 Ha: u < 20 for temp
```{r temp}
xbar = mean(forest_fires_mod$temp)
u = 20
s = sd(forest_fires_mod$RH)
n = nrow(forest_fires_mod)
t = (xbar-u)/(s/sqrt(n))
pval = pt(t, df=n-1) 
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", s/sqrt(n)), quote = FALSE)
print(paste0("t-score: ", t), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# H0: u = 20 Ha: u != 20 for area
```{r temp}
xbar = mean(forest_fires_mod$area)
u = 20
s = sd(forest_fires_mod$area)
n = nrow(forest_fires_mod)
t = (xbar-u)/(s/sqrt(n))
pval = 2*pt(t, df=n-1) 
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", s/sqrt(n)), quote = FALSE)
print(paste0("t-score: ", t), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# Machine Learning

## What are we trying to predict

Will a forest fire burn area level be large or small?

## Prepare the data set for the machine learning prediction task

```{r ml-setup}
forest_learning <- forest_fires_mod %>% select(-month,-X,-Y,-area,-log_area,-rain,-temp,-RH,-wind)
forest_learning %>% head()
```

## Prediction Algorithm

``` {r ml-alg}
library(caret)
set.seed(1234, sample.kind = "Rounding")

# create the cross-validation partition
cv_partition <- createFolds(forest_learning$burn_area_level,k=5)

# setup training parameters
fit_control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 8,
  #indexOut = cv_partition,
  summaryFunction=twoClassSummary,
  classProbs=TRUE,
  savePredictions=TRUE)

# a function to obtain performance data
# (tpr and fpr) over the given cross validation
# partitions, for the number of trees in the
# random forest
get_roc_data <- function(cv_partition, final_df) {
  mean_fpr <- seq(0, 1, len=100)
  aucs <- numeric(length(cv_partition))
  
  # iterate over folds
  res <- lapply(seq_along(cv_partition),  function(i) {
    # train the random forest 
    fit <- train(burn_area_level~(FFMC+DMC+DC+ISI)*season,
                        data = final_df[-cv_partition[[i]],], # all but the holdout set
                        method = "knn",
                        trControl = fit_control,
                        preProcess = c("center","scale"),
                        tuneLength = 20,
                        metric="ROC")
    
    # make predictions on the holdout set
    preds <- predict(fit, newdata=final_df[cv_partition[[i]],],type="prob")$large
    
    # compute tpr and fpr from the hold out set
    perf <- ROCR::prediction(preds, final_df$burn_area_level[cv_partition[[i]]]) %>%
      ROCR::performance(measure="tpr", x.measure="fpr")

    fpr <- unlist(perf@x.values)
    tpr <- unlist(perf@y.values)
    
    # interpolate the roc curve over 0, 1 range
    interp_tpr <- approxfun(fpr, tpr)(mean_fpr)
    interp_tpr[1] <- 0.0
    
    # collect values for this fold
    data_frame(fold=rep(i, length(mean_fpr)), fpr=mean_fpr, tpr=interp_tpr)
  })
  
  # combine values across all folds
  # into a single data frame
  do.call(rbind, res)
}

# calculate area under the ROC curve
# from tpr and fpr values across folds
compute_auc <- function(curve_df) {
  curve_df %>% 
    group_by(fold) %>%
    summarize(auc=pracma::trapz(fpr, tpr))
}
```

## Execute the prediction

``` {r ml-execution}
curve_df <- get_roc_data(cv_partition,forest_learning)
auc_df <- compute_auc(curve_df)
```

## Plot AUROC

```{r}
ggplot(auc_df, aes(x=fold, y=auc)) +
  geom_point()
```

## Plot ROC Curve

```{r}
curve_df %>%
  group_by(fpr) %>%
  summarize(tpr = mean(tpr)) %>%
  ggplot(aes(x=fpr, y=tpr)) +
    geom_line()
```

# Conclusion

# References