---
title: "Forest Fires NE Portugal"
author: "Shaurya Srivastava, Ishas Kekre, Anjali Viramgama"
date: "5/18/2020"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'docs/index.html')) })
output:
  html_document:
    theme: united
    highlight: breezedark
    number_sections: true
---

CMSC320 Final Project



```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(tidyverse)
library(broom)
library(gapminder)
library(stringr)

```
 
# Introduction
Every year, it seems that there is a disastrous fire breakout in different regions, and if we were able to predict it, it would be easy to control it and take precautions beforehand. In this page, we have tried to predict theburned area of forest fires, by using meteorological data in the Forest Fires Dataset provided by UCI's Machine Learning Repossitory.
# Forest Fires Information
Dataset characteristics: Multivariate
Attribute characteristics: Real
Number of Instances: 517
Number of Attributes: 13
Area: Physical

# Dataset Used

The dataset used is from http://archive.ics.uci.edu/ml/datasets/Forest+Fires. It contains data about forest fires that have occurred at Northeast Portugal.

# Understanding the Dataset
The dataset has the following attributes:
1. X - x-axis spatial coordinate 
2. Y - y-axis spatial coordinate
3. month - month of the year: 'jan' to 'dec'
4. day - day of the week: 'mon' to 'sun'
5. FFMC - FFMC index from the FWI system: 18.7 to 96.20
6. DMC - DMC index from the FWI system: 1.1 to 291.3
7. DC - DC index from the FWI system: 7.9 to 860.6
8. ISI - ISI index from the FWI system: 0.0 to 56.10
9. temp - temperature in Celsius degrees: 2.2 to 33.30
10. RH - relative humidity in %: 15.0 to 100
11. wind - wind speed in km/h: 0.40 to 9.40
12. rain - outside rain in mm/m2 : 0.0 to 6.4
13. area - the burned area of the forest (in ha): 0.00 to 1090.84
(this output variable is very skewed towards 0.0, thus it may make
sense to model with the logarithm transform).

Here is what different components of the FWI system mean:
FFMC- Fine Fuel Moisture Code, calculated using temperature, humidity, wind and rain.
DMC - Duff Moisture Code, calculated using temperature, relative humidity and rain.
DC - Drought Code, calculated using temperature and rain
ISI - Initial Spread Index, calculated using wind.
# Data Preparation

## Load the Dataset

```{r}
csv_file <- "forestfires.csv"
forest_fires <- read.csv(csv_file, stringsAsFactors = FALSE)
forest_fires %>% head()
```

## Modify Dataset as Needed 

Normally, this means dropping any columns that are unnecessary and adding columns that will be useful for further data analysis. We add a column with log of areas for better naalysis and convert all the months, which are from Jan to Dec to their respective numbers (1 to 12). We also add a season column (Fall, Spring, Summer and Winter) based on the months, which we will use for Hypothesis testing, and we de=ivide the burn area level into two parts, big and small, depending on it's area. 

### Add a column with the log of the areas. Drop the day column. Convert the month abbreviations to numbers

```{r}
forest_fires_mod <- forest_fires %>% 
  mutate(log_area = log(forest_fires$area + 1)) %>%
  select(-day) 
forest_fires_mod$month <- sapply(forest_fires_mod$month,function(x) grep(paste("(?i)",x,sep=""),month.abb))
```

### Add a column for the season

The season column is based on the month column. 

```{r}
forest_fires_mod <- forest_fires_mod %>% 
  mutate(season="winter")
forest_fires_mod$season[forest_fires_mod$month>2 & forest_fires_mod$month<6] <- "spring"
forest_fires_mod$season[forest_fires_mod$month>5 & forest_fires_mod$month<9] <- "summer"
forest_fires_mod$season[forest_fires_mod$month>8 & forest_fires_mod$month<12] <- "fall"
forest_fires_mod$season <- as.factor(forest_fires_mod$season)
```

### Add a column for the burn area level

There are two burn area levels, small and large. Burn area has a small level when log(area) <= 2 and a large level when log(area) > 2.

```{r}
forest_fires_mod <- forest_fires_mod %>% 
  mutate(burn_area_level="small")
forest_fires_mod$burn_area_level[forest_fires_mod$log_area>2] <- "large"
forest_fires_mod$burn_area_level <- as.factor(forest_fires_mod$burn_area_level)
```

### Display the first few rows of the new modified dataframe.

```{r} 
forest_fires_mod %>% head()
```

# Exploratory Data Analysis
We first try to find if there is any correaltion between density and months.

``` {r hist}
forest_fires_mod %>%
  ggplot(mapping=aes(x=month),  breaks=100) +
    ggtitle("                                                   Histogram of Months") +
    geom_histogram(aes(y=..density..),binwidth=1.5) +
    geom_density(alpha=.2, fill="#FF6666") +
    scale_x_continuous(name ="Months", 
                    breaks=seq(1,12,1))
```

There seems to be a higher density between months July to October, i.e. in summer and fall. To further confirm this, we plot a pie chat showing the % of total fires per season.
``` {r pie}
t = nrow(forest_fires_mod)
s = nrow(forest_fires_mod %>% filter(season == 'summer'))
sp = nrow(forest_fires_mod %>% filter(season == 'spring'))
f = nrow(forest_fires_mod %>% filter(season == 'fall'))
w = nrow(forest_fires_mod %>% filter(season == 'winter'))
seasons <- data.frame(
  group = c("Summer", "Spring", "Fall", "Winter"),
  value = c(s/t, sp/t, f/t,w/t)
  )
seasons %>%
  ggplot(aes(x="", y=value, fill=group)) + geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) + geom_text(aes(label = paste0(round(value*100), "%")), position = position_stack(vjust = 0.5)) +
  labs(x = NULL, y = NULL, fill = NULL, title = "                             Seasonal Fires") +
  theme_classic() + theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())
```

This clearly shows that the tendency of a forest fire breaking out in Summer and Fall is higher than the other 2 seasons. 
We plot another graph to show what areas of the forest are most affected by forest fires, in X and Y coordinates.
``` {r map}

forest_fires_mod %>% 
  ggplot(aes(x=X, y=Y) ) +
  ggtitle("                                           Forest Fire Map") +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  scale_x_continuous(name ="X-coord", 
                    breaks=seq(1,9,1)) +
  scale_y_continuous(name ="Y-coord", 
                    breaks=seq(2,9,1))
```

# Hypothesis Testing
For our hypothesis testing, we assumed that the probability of a forest fire occuring in every season is equal, i.e. it is 0.25 for summer, spring, winter and fall.
We test it out for Fall specifically, assuming p = 0.25
# H0: p = 0.25 Ha: p > 0.25 for fall
```{r fall}
fall_sample <- forest_fires_mod %>% filter(season == 'fall')
xbar = nrow(fall_sample)/nrow(forest_fires_mod)
u = 0.25
sigma = 0.25*(1-0.25)/nrow(forest_fires_mod)
z = (xbar-u)/(sqrt(sigma))
pval = pnorm(z, lower.tail=FALSE)
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", sqrt(sigma)), quote = FALSE)
print(paste0("z-score: ", z), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# H0: u = 20 Ha: u < 20 for temp
```{r temp}
xbar = mean(forest_fires_mod$temp)
u = 20
s = sd(forest_fires_mod$RH)
n = nrow(forest_fires_mod)
t = (xbar-u)/(s/sqrt(n))
pval = pt(t, df=n-1) 
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", s/sqrt(n)), quote = FALSE)
print(paste0("t-score: ", t), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# H0: u = 20 Ha: u != 20 for area
```{r temp}
xbar = mean(forest_fires_mod$area)
u = 20
s = sd(forest_fires_mod$area)
n = nrow(forest_fires_mod)
t = (xbar-u)/(s/sqrt(n))
pval = 2*pt(t, df=n-1) 
print(paste0("xbar: ", xbar), quote = FALSE)
print(paste0("mean: ", u), quote = FALSE)
print(paste0("std: ", s/sqrt(n)), quote = FALSE)
print(paste0("t-score: ", t), quote = FALSE)
print(paste0("p-value: ", pval), quote = FALSE)
```
# Machine Learning

## What are we trying to predict

Will a forest fire burn area level be large or small?

## Prepare the data set for the machine learning prediction task

```{r ml-setup}
forest_learning <- forest_fires_mod %>% select(-month,-X,-Y,-area,-log_area,-rain,-temp,-RH,-wind)
forest_learning %>% head()
```

## Prediction Algorithm

``` {r ml-alg}
library(caret)
set.seed(1234, sample.kind = "Rounding")

# create the cross-validation partition
cv_partition <- createFolds(forest_learning$burn_area_level,k=5)

# setup training parameters
fit_control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 8,
  #indexOut = cv_partition,
  summaryFunction=twoClassSummary,
  classProbs=TRUE,
  savePredictions=TRUE)

# a function to obtain performance data
# (tpr and fpr) over the given cross validation
# partitions, for the number of trees in the
# random forest
get_roc_data <- function(cv_partition, final_df) {
  mean_fpr <- seq(0, 1, len=100)
  aucs <- numeric(length(cv_partition))
  
  # iterate over folds
  res <- lapply(seq_along(cv_partition),  function(i) {
    # train the random forest 
    fit <- train(burn_area_level~(FFMC+DMC+DC+ISI)*season,
                        data = final_df[-cv_partition[[i]],], # all but the holdout set
                        method = "knn",
                        trControl = fit_control,
                        preProcess = c("center","scale"),
                        tuneLength = 20,
                        metric="ROC")
    
    # make predictions on the holdout set
    preds <- predict(fit, newdata=final_df[cv_partition[[i]],],type="prob")$large
    
    # compute tpr and fpr from the hold out set
    perf <- ROCR::prediction(preds, final_df$burn_area_level[cv_partition[[i]]]) %>%
      ROCR::performance(measure="tpr", x.measure="fpr")

    fpr <- unlist(perf@x.values)
    tpr <- unlist(perf@y.values)
    
    # interpolate the roc curve over 0, 1 range
    interp_tpr <- approxfun(fpr, tpr)(mean_fpr)
    interp_tpr[1] <- 0.0
    
    # collect values for this fold
    data_frame(fold=rep(i, length(mean_fpr)), fpr=mean_fpr, tpr=interp_tpr)
  })
  
  # combine values across all folds
  # into a single data frame
  do.call(rbind, res)
}

# calculate area under the ROC curve
# from tpr and fpr values across folds
compute_auc <- function(curve_df) {
  curve_df %>% 
    group_by(fold) %>%
    summarize(auc=pracma::trapz(fpr, tpr))
}
```

## Execute the prediction

``` {r ml-execution}
curve_df <- get_roc_data(cv_partition,forest_learning)
auc_df <- compute_auc(curve_df)
```

## Plot AUROC

```{r}
ggplot(auc_df, aes(x=fold, y=auc)) +
  geom_point()
```

## Plot ROC Curve

```{r}
curve_df %>%
  group_by(fpr) %>%
  summarize(tpr = mean(tpr)) %>%
  ggplot(aes(x=fpr, y=tpr)) +
    geom_line()
```

# Conclusion
We conclude using ML,we can predict the area that will be affected by forest fires. We also conclude that the fires are slightly dependent on the season and the month.
# References